\begin{abstract}
Initial Abstract by Claude: Deep learning models deployed in safety-critical settings can fail silently under distribution shift, producing confident but unreliable outputs with no mechanism for self-diagnosis. We study this through seismic phase picking, showing that PhaseNet's P-wave detection rate drops from 85.5\% to 35.1\% between California and Kazakhstan while confidence distributions remain nearly identical, rendering standard confidence-based deferral ineffective. We develop a regime-aware deferral framework that combines an external Hidden Markov Model tracking the seismic environment with Beta-Bernoulli reliability estimation, producing a deferral score that integrates environmental uncertainty with model confidence. The regime-aware score concentrates 90.9\% of deferrals during seismically active periods (vs.\ 36.4\% for confidence-only), directing scarce human analyst time to the moments of greatest environmental uncertainty---though confidence-only deferral achieves higher raw system accuracy at matched deferral rates. The framework demonstrates a general principle for safety-critical ML: when model self-assessment fails under distribution shift, structured external context can provide the reliability signal that internal uncertainty cannot.

Slightly Tweaked Abstract: 
Deep learning models deployed in safety-critical settings can fail silently under distribution shift, producing confident but unreliable outputs with no way to self-diagnose the failure. We study this in seismic phase picking, the task of identifying earthquake wave arrivals (picks) in seismograms. PhaseNet, a widely-deployed deep learning model trained on Northern California data, sees its primary (P) wave detection rate collapse from 83.9\% to 35.1\% when deployed to Kazakhstan, a seismically active but data-sparse region where automated tools are most needed yet least validated. PhaseNet's own confidence scores are nearly indistinguishable between the two regions (AUROC of 0.699 in California vs 0.701 in Kazakhstan), so a system monitoring confidence alone would have no way of knowing that detection performance had collapsed. The core question for an earthquake monitoring agency then becomes: when should the system trust its automated picks, and when should it route them to an analyst for review? We develop a regime-aware deferral framework to answer this, combining an external Hidden Markov Model of the seismic environment with Beta-Bernoulli estimation of per-regime model reliability. The regime-aware score defers 90.9\% of picks made during seismically active periods, compared to 36.4\% when deferring on PhaseNet's confidence alone, directing scarce analyst time toward regimes where reliability rests on the fewest observations, at a marginal cost in accuracy (0.619 vs 0.651). This matters most in data-sparse regions where active seismic events are rare and reliable detection is most critical. The regions that stand to benefit most from automation are precisely those where automated tools are least validated, making frameworks that know when not to trust the model essential for equitable deployment.


\end{abstract}
