\section{Data Collection and Preprocessing}
\label{app:data}

% TODO: Populate with full details from notebook Steps 1-2
% A.1 Kazakhstan station selection: metadata table, filtering pipeline, station map
% A.2 Event catalog: ISC query parameters, magnitude distribution, M4.5+ proximity
% A.3 California baseline: NCEDC query, event selection, per-event waveform counts
% A.4 Per-station breakdowns: detection rate, precision, confidence per station

Details of the data collection and preprocessing pipeline will be provided here, including the full station metadata table, filtering criteria, and per-station performance breakdowns.

\section{HMM Specification and Sensitivity Analysis}
\label{app:hmm}

% TODO: Populate from notebook Step 5
% B.1 Parameter derivation: transition matrix, station-specific Poisson rates, emission verification
% B.2 Forward algorithm: full specification, hard injection handling
% B.3 Sensitivity analysis: recovery times under 6 perturbations × 4 station-event pairs

The full HMM parameter specification, forward algorithm details, and sensitivity analysis results will be provided here.

\section{Regime Posterior Visualisations}
\label{app:posteriors}

% TODO: Generate from notebook Step 5c
% Stacked area charts for all 11 stations, grouped by triggered vs untriggered

The complete set of regime posterior plots for all 11 stations will be provided here.

% YASMIN: Populated from Notebook - 3 March, Steps 6-7
\section{Deferral Framework Details}
\label{app:deferral}

\subsection{Beta-Bernoulli Convergence}

% TODO: Figure — Beta-Bernoulli posterior evolution plots from Step 6b
% Show alpha, beta, theta, sigma evolution over 190 picks for each regime

The Beta-Bernoulli model is initialised with $\text{Beta}(2, 2)$ priors for all three regimes and updated sequentially over the 190 Kazakhstan picks using soft regime assignment. \Cref{tab:beta_final} reports the final posterior parameters.

\begin{table}[h]
    \centering
    \caption{Beta-Bernoulli posterior parameters after processing all 190 Kazakhstan picks.}
    \label{tab:beta_final}
    \small
    \begin{tabular}{@{}lcccc@{}}
        \toprule
        Regime & $E[\theta_r]$ & $\sigma_r$ & Eff.\ samples & Interpretation \\
        \midrule
        Quiet & 0.425 & 0.038 & 164.6 & Converged \\
        Active & 0.600 & 0.123 & 11.0 & High uncertainty \\
        Decaying & 0.400 & 0.111 & 14.4 & High uncertainty \\
        \bottomrule
    \end{tabular}
\end{table}

The Quiet regime accumulates the vast majority of observations (164.6 effective samples), yielding a narrow posterior ($\sigma_Q = 0.038$). Active and Decaying regimes remain data-sparse (11.0 and 14.4 effective samples respectively), with posterior standard deviations approximately $3\times$ larger. This asymmetry drives the deferral score: $\sigma_t$ increases during Active and Decaying periods, suppressing confidence and triggering deferral.

\subsection{Threshold Sensitivity}

% TODO: Figure — Full threshold sweep from Step 6c
% System accuracy vs deferral rate (0-50\%) for both methods

At the 25\% operating point, confidence-only defers the 48 picks with lowest PhaseNet confidence, achieving 0.651 system accuracy. The combined score defers the 48 picks with the lowest regime-aware scores, achieving 0.619. Confidence-only maintains a slight accuracy advantage across most deferral rates because it directly targets the picks most likely to be incorrect, whereas the combined score prioritises regime-uncertain picks regardless of individual confidence.

\subsection{Regime-Stratified Deferral Rates}

\begin{table}[h]
    \centering
    \caption{Regime-stratified deferral rates at 25\% overall deferral.}
    \label{tab:regime_deferral_detail}
    \small
    \begin{tabular}{@{}lccc@{}}
        \toprule
        Regime & Picks & Conf-only deferred & Combined deferred \\
        \midrule
        Quiet ($\pi_Q > 0.9$) & 155 & 36 (23.2\%) & 27 (17.4\%) \\
        Active ($\pi_A > 0.5$) & 11 & 4 (36.4\%) & 10 (90.9\%) \\
        Decaying ($\pi_D > 0.5$) & 15 & 5 (33.3\%) & 9 (60.0\%) \\
        Mixed & 9 & 3 (33.3\%) & 2 (22.2\%) \\
        \bottomrule
    \end{tabular}
\end{table}

\subsection{Human-in-the-Loop Capacity Curves}

% TODO: Figure — Capacity curves from Step 7a
% System accuracy vs analyst capacity (0-50\%) for both methods

\begin{table}[h]
    \centering
    \caption{System accuracy at different analyst capacity levels (48 picks deferred).}
    \label{tab:capacity}
    \small
    \begin{tabular}{@{}lccc@{}}
        \toprule
        Capacity & Conf-only & Combined & Utilisation \\
        \midrule
        10\% & 0.527 & 0.485 & 100\% \\
        15\% & 0.572 & 0.519 & 100\% \\
        20\% & 0.606 & 0.569 & 100\% \\
        25\% & 0.651 & 0.619 & 100\% \\
        30\% & 0.651 & 0.619 & 84.2\% \\
        \bottomrule
    \end{tabular}
\end{table}

Both methods saturate at 25\% analyst capacity, at which point all 48 deferred picks have been reviewed. Below saturation, the combined score directs 20.8\% of analyst capacity to Active picks and 18.8\% to Decaying, compared with 8.3\% and 10.4\% for confidence-only.

\subsection{Analyst Accuracy Sensitivity}

% TODO: Figure — Analyst sensitivity sweep from Step 7c
% System accuracy vs analyst accuracy (70-99\%) for both methods + baseline

\begin{table}[h]
    \centering
    \caption{System accuracy as a function of assumed analyst accuracy.}
    \label{tab:analyst_sensitivity}
    \small
    \begin{tabular}{@{}lccc@{}}
        \toprule
        Analyst accuracy & No deferral & Conf-only & Combined \\
        \midrule
        75\% & 0.432 & 0.600 & 0.568 \\
        85\% & 0.432 & 0.625 & 0.594 \\
        95\% & 0.432 & 0.651 & 0.619 \\
        \bottomrule
    \end{tabular}
\end{table}

Confidence-only deferral exceeds the no-deferral baseline when analyst accuracy exceeds 8.3\%; the combined score exceeds it when analyst accuracy exceeds 20.8\%. The higher breakeven for the combined score reflects the fact that it defers some high-confidence picks during uncertain regimes---picks that may already be correct---meaning the analyst must be reasonably accurate to improve on the automated decision.
