\section{Background and Related Work}
\label{sec:background}

\subsection{PhaseNet and Deep Learning for Seismology}

% One paragraph. What PhaseNet is (Zhu & Beroza 2019): U-Net CNN, three-component waveforms → P/S probability traces + confidence.
% Mention SeisBench (Woollam et al. 2022) for evaluation interface.
% Note the trend: models trained on well-instrumented networks deployed globally → distribution shift.
% Use Sera's description of the architecture but keep it concise.

PhaseNet \citep{zhu2019phasenet} is a U-Net convolutional neural network that takes three-component seismic waveforms (recordings of ground motion in the vertical, north-south, and east-west directions) as input and outputs a probability at each timestep for P-wave, S-wave, and noise classes. Picks are extracted by identifying peaks in the per-timestep probabilities and retaining those that exceed a configurable confidence threshold. PhaseNet was trained on approximately 780{,}000 waveforms from the Northern California Seismic Network and has become one of the most widely-used tools for automated phase picking, accessible through the standardised SeisBench interface \citep{woollam2022seisbench}. It is increasingly deployed to data-sparse regions globally, the distribution shift scenario we study here.

\subsection{Distribution Shift and the Limits of Model Self-Assessment}
\label{sec:bg_shift}

% Two paragraphs.
% Para 1: Neural network miscalibration (Guo et al. 2017), degradation under shift (Ovadia et al. 2019).
% Core module concept: epistemic uncertainty should increase OOD → natural deferral signal.
% We test this for PhaseNet and find it fails.
% Para 2: Selective prediction history (Chow 1970, Geifman & El-Yaniv 2017), learning to defer (Madras et al. 2018, Mozannar & Sontag 2020).
% Key distinction: most assume (a) informative uncertainty or (b) labelled deployment data. Neither holds here.
% Our gap: external environmental context instead of model self-assessment.

Neural network confidence scores are often poorly calibrated even on in-distribution data \citep{guo2017calibration}, and calibration degrades further under distribution shift \citep{ovadia2019uncertainty}. A core principle of uncertainty quantification is that epistemic uncertainty should increase in unfamiliar regions of input space, providing a natural signal for selective prediction or deferral. We test whether this holds for PhaseNet and find that it does not: confidence scores carry essentially no information about whether the model is operating in- or out-of-distribution, with P-wave AUROC virtually identical between California (0.699) and Kazakhstan (0.701). This motivates the search for an external reliability signal.

When model confidence is informative, the standard response is to let the model abstain on uncertain predictions \citep{chow1970optimum,geifman2017selective}. Learning-to-defer frameworks \citep{madras2018predict,mozannar2020consistent} extend this idea by optimising the decision of when to defer to a human expert, balancing automated accuracy against expert review cost. These frameworks typically assume either that model uncertainty is informative or that a learned rejector can be trained on labelled deployment-domain data. In our setting, neither assumption holds. PhaseNet's confidence is uninformative under distribution shift (\cref{sec:confidence}), and labelled Kazakhstan data is scarce. Our framework avoids both requirements by using external environmental context to drive deferral decisions.

\subsection{HMMs and Seismic Regime Modelling}
\label{sec:bg_hmm}

% One paragraph. Omori's law (1894), ETAS (Ogata 1988), the Q→A→D cycle.
% HMM is domain-informed, not arbitrary. Connects to module emphasis on incorporating domain knowledge.

The seismic cycle of quiet background $\to$ mainshock $\to$ aftershock decay is well-established in seismology \citep{omori1894aftershocks, ogata1988etas}. Aftershock rates follow Omori's law, decaying as a power law with time since the mainshock \citep{utsu1995centenary}. This Quiet $\to$ Active $\to$ Decaying structure maps naturally onto Hidden Markov Model assumptions \citep{rabiner1989hmm}: the hidden states represent the underlying seismic regime, which cannot be observed directly, but generates observable weekly earthquake counts, with transition probabilities encoding how long each regime typically lasts before shifting to the next. The HMM encodes this established seismological structure directly, using domain knowledge to define the hidden states and their transitions rather than learning them from data.

Poisson HMMs have been used to model weekly earthquake counts in seismically active regions, where they detect shifts between quiet and active periods that simpler models miss. \citet{orfanogiannaki2010poisson} apply a Poisson HMM to weekly counts in Greece, demonstrating that the model detects regime shifts that traditional rate-change analysis misses entirely, supporting the broader principle that state-based models capture temporal structure that simpler threshold methods do not. \citet{felix2022poisson} independently reach the same modelling choice for Metro Manila, further demonstrating that Poisson HMMs capture earthquake activity patterns that simpler approaches do not.

\citet{orfanogiannaki2010poisson} and \citet{felix2022poisson} both learn their HMM parameters from data, but with only five triggering events in Kazakhstan, data-driven parameter estimation would yield unreliable posteriors \citep{naylor2023bayesian}. We instead set transition probabilities directly from established Omori-law aftershock timescales \citep{utsu1995centenary, reasenberg1989earthquake}, following the principle that domain-knowledge parameterisation is preferable to fitting parameters purely from data when data is sparse \citep{naylor2023bayesian}. \citet{beyreuther2012hmm} take a similar approach in HMM-based earthquake detection with only 21 training events, constraining model parameters rather than learning them freely.

The most established alternative to HMMs for modelling aftershock sequences is ETAS \citep{ogata1988etas}, which models seismicity as a self-exciting process with Omori-law triggering rates. We considered ETAS but it requires more data than our setting provides. Reliable ETAS inversion requires both quiet periods and active sequences in the training data, and is sensitive to short-term catalog incompleteness after large events \citep{naylor2023bayesian}, which are requirements our dataset of five triggering events cannot satisfy. This practical limitation is well-documented: \citet{touati2016seismicity} identify catalog incompleteness as a fundamental bias in ETAS parameter estimation after large events. HMM avoids these requirements while retaining the property we need: a sequential, probabilistic estimate of the current seismic regime that updates as new observations arrive.

\subsection{Bayesian Reliability Tracking}
\label{sec:bg_bayes}

Beta-Bernoulli models provide a conjugate framework for tracking binary outcome reliability with uncertainty \citep{murphy2012machine}: a Beta prior over success probability is updated sequentially as outcomes arrive, yielding a posterior that naturally widens when data are sparse. This is well-suited to our setting, where PhaseNet's pick-level correctness is a binary outcome and regime-stratified sample sizes are highly unequal: 155 Quiet picks versus 11 Active and 15 Decaying. The posterior width over per-regime precision enters directly into our deferral score, so that data-sparse regimes are penalised not just for lower point-estimate reliability but for higher uncertainty about that reliability. Expected Calibration Error \citep{guo2017calibration} and AUROC \citep{hanley1982roc} are used to evaluate whether PhaseNet's native confidence scores are sufficient for deferral before introducing the regime-aware component.


