\section{Discussion and Limitations}
\label{sec:discussion}

\subsection{Limitations}
\label{sec:limitations}

\paragraph{Small regime-stratified samples.} The Active and Decaying performance analysis relies on a single station (PDGK) with 11--13 picks per regime. We frame regime-stratified results as suggestive, not statistically conclusive. A longer study period or a region with more frequent large earthquakes would strengthen this evidence.

\paragraph{Single target region.} We demonstrate the framework on one distribution shift (California $\to$ Kazakhstan). The architecture generalises, but the HMM parameters and deferral threshold would need recalibration for each new deployment region.

\paragraph{Small evaluation dataset.} The entire Kazakhstan evaluation is based on 190 PhaseNet picks against 202 P-wave and 74 S-wave ISC arrivals. This sample size limits the statistical power of finer-grained analyses such as per-regime accuracy estimates.

\paragraph{Simulated analyst.} The human-in-the-loop evaluation uses a synthetic analyst with fixed accuracy (95\%) and deterministic prioritisation. The sensitivity analysis (\cref{sec:hitl}) shows both methods exceed the no-deferral baseline across all plausible analyst accuracies, but a real deployment would involve variable analyst attention, fatigue, and disagreement between analysts. The simulation provides a controlled comparison between deferral methods but does not capture the operational complexity of real-time analyst review.

\paragraph{Catalog dependence.} Regime triggers require a functioning earthquake catalog. In the most data-sparse regions, even the catalog may be incomplete, potentially missing events that should trigger Active. This creates a floor on how well the framework can perform in truly unmonitored areas, and is the primary constraint on transferability to other regions (see Broader Impact).

\paragraph{Weekly resolution.} The HMM operates at weekly granularity, which may be too coarse to capture rapid regime transitions in the first hours after a mainshock. Sub-weekly resolution would require denser seismicity observations.

\paragraph{Accuracy vs.\ targeting trade-off.} The combined score achieves lower raw system accuracy than confidence-only at matched deferral rates. Practitioners who prioritise accuracy over temporal targeting may prefer confidence-only deferral.

\paragraph{Recall vs.\ precision failure modes.} As identified in this study, PhaseNet's dominant failure mode under distribution shift is missed detections (recall failure) rather than incorrect picks (precision failure). The Beta-Bernoulli model tracks precision: whether picks that PhaseNet \emph{makes} are correct. It cannot directly detect picks that PhaseNet \emph{fails to make}, since missed detections have no timestamp to align with regime posteriors. This is a fundamental limitation of any pick-level deferral system when applied to recall-dominated failure modes.

\paragraph{No formal safety guarantees.} The framework provides empirical deferral targeting but no theoretical bounds on error rates or coverage. The deferral threshold $\tau$ is set to achieve a target deferral rate, not a target error rate or safety level. Methods such as conformal prediction can provide finite-sample coverage guarantees, and integrating such guarantees into the deferral decision is a natural direction for future work.

\paragraph{Discrete regime assumption.} The HMM assumes three discrete states (Quiet, Active, Decaying). The real seismic environment is continuous. Intermediate states, overlapping aftershock sequences from multiple mainshocks, or regime dynamics that do not follow Markov assumptions (e.g., long-memory processes) are not captured by this model.

\paragraph{Exchangeability within regimes.} The Beta-Bernoulli model treats all picks within a regime as draws from the same distribution, ignoring within-regime heterogeneity. Station-specific effects, magnitude-dependent performance, and signal-to-noise variation could all cause two picks during the same Active week to have very different reliability. The model assigns them the same $\sigma_t$ regardless.

\paragraph{Fixed HMM parameters.} The transition matrix and emission rates are set once from domain knowledge and do not adapt over time. Over longer deployment timescales, seismicity patterns could change (e.g., activation of a previously quiet fault zone), and the system would not detect this without manual re-estimation of parameters.

\paragraph{Ground truth quality.} The ISC Reviewed Bulletin is treated as ground truth, but ISC picks have their own measurement errors and the bulletin's completeness varies by region. Evaluation metrics are therefore estimates relative to the best available human reference, not absolute measures of performance.

\paragraph{Baseline comparisons.} We compare against confidence-only deferral and no deferral, which are the approaches a practitioner would try first. We do not compare against more sophisticated methods such as ensemble disagreement or learned rejectors because these require either architectural modifications to PhaseNet or labelled deployment-domain data, neither of which is available in our setting.

\subsection{Do the Experiments Support the Claims?}
\label{sec:self_eval}

We cannot claim the regime signal \emph{causes} better deferral---the comparison is observational, not ablated in a controlled experiment. What we show is that the two methods defer systematically different picks, and that the combined score's pattern aligns with operationally relevant regime structure.

The baselines are appropriate: confidence-only deferral is the standard approach a practitioner would use, and it is a strong baseline that wins on raw accuracy. We do not compare against more sophisticated methods (e.g., learned rejectors, ensemble disagreement) because these require either labelled deployment data or architectural modifications to PhaseNet, neither of which is available in our setting.

System accuracy captures overall correctness; regime-stratified deferral rates capture temporal targeting. We report both rather than selecting the metric that favours our method.

\subsection{Broader Implications for Safety-Critical ML}
\label{sec:implications}

This work illustrates a general principle: when a model cannot self-diagnose failure under distribution shift, \emph{external context about the deployment environment} can provide the reliability signal that internal uncertainty cannot. The specific context here is seismic regime, but the architecture applies wherever deployment conditions are structured and observable, these conditions affect model reliability systematically, and the model's own uncertainty fails to reflect this---for example, medical imaging across hospitals with different equipment profiles, autonomous vehicles across geographies with different road conditions, or environmental monitoring systems deployed across ecosystems.
