\section{Introduction}
\label{sec:intro}

Safety-critical machine learning systems are increasingly deployed under distribution shift: a model trained in one setting is used in another, and its performance may degrade without warning. The standard defence is to use the model's own uncertainty estimates to flag unreliable predictions, deferring low-confidence outputs to a human expert. This assumes the model knows what it does not know. When this assumption fails, the system produces confident but incorrect outputs with no internal signal to trigger human review.

We study this failure concretely through seismic phase picking, the task of identifying earthquake wave arrivals in continuous seismograms. PhaseNet \citep{zhu2019phasenet}, trained on Northern California data, sees its P-wave detection rate collapse from 83.9\% to 35.1\% when deployed to Kazakhstan. Despite this collapse, mean confidence for matched picks is nearly unchanged (0.749 vs.\ 0.747), and P-wave AUROC for correct/incorrect discrimination is virtually identical across regions (0.699 vs.\ 0.701). Confidence-based deferral has no signal that anything has changed.

We propose a regime-aware deferral framework that replaces model self-assessment with external environmental context. An HMM tracks the seismic environment (quiet background, active aftershock sequences, gradual decay) and a Beta-Bernoulli model estimates PhaseNet's per-regime reliability. The combined deferral score weights PhaseNet's confidence against regime uncertainty, routing picks from poorly-observed regimes to a human analyst. Our contributions are:

\begin{enumerate}
    \item An empirical demonstration that PhaseNet's confidence scores are uninformative under distribution shift: the model fails silently, with no internal signal to trigger human review. This establishes that standard uncertainty quantification tools are insufficient for deferral in this setting and motivates the need for an external reliability signal.
    \item A regime-aware monitoring framework using an HMM to track the seismic environment, providing a structured external signal that model confidence cannot. This enables deferral decisions that reflect when the deployment environment is most uncertain, rather than when individual picks are most uncertain.
    \item A Bayesian deferral mechanism combining regime uncertainty with pick-level confidence, which concentrates deferrals during seismically active periods rather than spreading them uniformly across time. This directs scarce analyst time toward the periods of greatest environmental disturbance, with honest evaluation showing that confidence-only deferral achieves higher raw system accuracy at matched deferral rates.
\end{enumerate}

The remainder of this paper is organised as follows. Section~\ref{sec:background} reviews related work. Section~\ref{sec:method} presents the framework. Section~\ref{sec:setup} describes the experimental setup. Section~\ref{sec:results} presents results. Section~\ref{sec:discussion} discusses limitations and broader implications.

\section{Related Work}
\label{sec:background}

\subsection{Deep Learning Phase Picking and Distribution Shift}
\label{sec:bg_picking}

PhaseNet \citep{zhu2019phasenet} is a U-Net convolutional neural network trained on approximately 780{,}000 waveforms from the Northern California Seismic Network, producing continuous P-wave, S-wave, and noise probability traces. It is widely deployed through the standardised SeisBench interface \citep{woollam2022seisbench}. Models trained on well-instrumented networks are known to degrade when deployed to seismically distinct regions \citep{mousavi2020earthquake}, and neural network confidence scores are poorly calibrated under distribution shift \citep{ovadia2019uncertainty}, meaning the model does not signal when it is failing. We confirm both findings for PhaseNet in Kazakhstan (Section~\ref{sec:results}).

\subsection{Selective Prediction and Learning to Defer}
\label{sec:bg_defer}

% Selective prediction \citep{chow1970optimum, geifman2017selective} and the learning-to-defer framework \citep{madras2018predict, mozannar2020consistent} formalise when an automated system should route decisions to a human expert. Most approaches assume either that model uncertainty is informative or that a learned rejector can be trained on labelled deployment-domain data. Neither holds in our setting: PhaseNet's confidence is uninformative under shift, and labelled Kazakhstan data is scarce. Our framework addresses this gap by using external environmental context as the deferral signal.

Selective prediction \citep{chow1970optimum, geifman2017selective} assumes model confidence is informative enough to identify and defer unreliable predictions. The learning-to-defer framework \citep{madras2018predict, mozannar2020consistent} relaxes this by training a learned rejector, but requires labelled deployment-domain data to do so. Neither holds in our setting: PhaseNet's confidence is uninformative under shift, and labelled Kazakhstan data is scarce. Our framework addresses this gap by using external environmental context as the deferral signal.

\subsection{Seismic Regime Modelling}
\label{sec:bg_regime}

Aftershock sequences follow Omori's law \citep{omori1894aftershocks, utsu1995centenary}, decaying as a power law from a mainshock through elevated activity back to background seismicity. ETAS \citep{ogata1988etas} models this self-exciting process in detail, but reliable parameter estimation requires both active sequences and periods of quiescence \citep{naylor2023bayesian}, a requirement our five triggering events cannot meet, a limitation independently encountered by \citet{felix2022poisson} and \citet{touati2016seismicity}. Poisson HMMs offer a practical alternative: \citet{orfanogiannaki2010poisson} demonstrate that a Poisson HMM on weekly earthquake counts detects regime shifts that traditional rate-change analysis misses, and \citet{felix2022poisson} independently reach the same modelling choice for a different region. Both learn parameters from data via
expectation-maximisation. While transition probabilities can in principle be estimated from catalog data, our five triggering events provide too few observed regime transitions for reliable estimation. We instead set transition probabilities from domain knowledge, motivated by established aftershock statistics (Utsu et al., 1995).

% Extra notes: Both learn emission rates, transition probabilities, and state structure from data via expectation-maximisation; we fix the state structure from seismological domain knowledge and set transition probabilities from aftershock statistics, estimating only the quiet-period emission rate λQ from observed counts.